Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-06-01 article/main.tex
conclusion + numpy + stuff

\section{Introduction} %ce qu'on avait écrit pour le projet, peut-être pas incroyable...

@@

%\todo{Ca, c'est du blabla...}

@@

Not all pseudo-random generators are of cryptographic strength. In some
applications, it is not necessarily necessary \todo{lourd}: to be used in Monte-Carlo
numerical simulations or generate random choices in games, a relaxed,
non-cryptographic notion of pseudo-randomness may be sufficient. This allows for
faster algorithms. For instance, \textsf{python} standard library's
\texttt{random} module uses the Mersenne Twister~\cite{MatsumotoN98}. The
\textsf{C} library that comes along \texttt{gcc} (the \texttt{glibc}) uses a
(poor) truncated linear congruential generator by default to implement the
\texttt{rand} function.

@@

In any case, designers of conventional pseudo-random generators try to obtain
the simplest and fastest algorithm that passes the day's favourite test suite. A
few selected ones are shown in Fig.~\ref{fig:prngs}. \texttt{Lehmer64} is a
truncated linear congruential generator, touted as ``the fastest PRNG that
passes \texttt{BigCrush}''~\cite{LemireLehmer}. \texttt{Xorshift128} is a clever
implementation of a 128-bit LFSR with period $2^{128} - 1$ due to
Marsaglia~\cite{Marsaglia03}, using only a few simple 32-bit
operations. \texttt{Xorshift128+} is a improved version due to
Vigna~\cite{Vigna17} that returns the sum of two consecutive outputs of a
Xorshift LFSR; it passes the ``BigCrush'' test suite and is the default PRNG in
many Javascript implementations, including that in Google's V8 engine (Chrome),
Firefox and Safari.

@@

\textsf{PCG} stands for ``Permuted Congruential Generator'': it essentially
consists in applying a non-linear filtering function on top of a linear
congruential generator (in a way reminiscent to the venerable filtered
LFSRs). The resulting combination is fast and passes current statistical test
suites. The \textsf{PCG} family contains many members, but we focus on the
strongest one, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has a
128-bit internal state and produces 64 bits when clocked. It is the default
pseudo-random number generator in the popular \textsf{NumPy} scientific
computing package for \textsf{Python}.

@@

When the increment is unknown, things are a bit more complicated. This is the
default situation in \textsf{NumPy}, where both the state and the increment are
initialised using an external source of entropy. In this case, our prediction
algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
55 bits, then for each guess it solves an instance of CVP in dimension 4 (using
about 75 arithmetic operations). This recovers 64 more bits of information about
the difference between two successive states, and this is enough to filter the
bad guesses. This information can then be used in a subsequent and comparably
inexpensive phase to recover the entire internal state. On average, the whole
process requires a bit less than 20 000 CPU hours to complete.

We implemented the algorithms in \textsf{C}, then asked the designer of the PCG
family to send us ``challenge'' pseudo-random streams ; we ran our code and
emailed back the seeds used to generate the challenge streams the next day.\todo{et on avait tout juste, enfin j'espère}

@@

Knuth considered the problem of recovering the internal state of a truncated
linear congruential generator $x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$
are unknown~\cite{Knuth85}; he gave an algorithm exponential in the number of
truncated bits.

In 1989, J. Boyar studied further the problem \cite{Boyar1989}. She presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.

In 1988, A. Frieze \textit{et al} \cite{Frieze} improved the efficiency of reconstruction algorithms in simpler cases. They supposed that the multiplier $a$ and the modulus $2^k$ were known and used lattice-based technique to recover the LCG more efficiently.

Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment $c$ is zero) \cite{Lehmer}.

@@

\textsf{PCG64} has an internal state of 128-bit, which operate as a linear
congruential generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]
Where \todo{il manque une lettre, je suppose que c'est a} is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
the generator. The increment $c$ may be specified by the user of the PRNG to
produce different output streams with the same seed (just as the IV acts in a
stream cipher). If no value of $c$ is specified, then a default increment is
provided. Note that $c$ must be odd. The default values are:
\begin{align*}
  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
\end{align*}

@@

In the rest of this paper, we often perform arithmetic operations on integers
where only some bits are known. This leads to generation of unknown carries. If
$a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there is a
\emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
$0 \leq 1 \leq \beta$ \todo{je pense pas que ce soit ça}) such that:
\begin{align}
  (a \boxplus b)[i:j]  &= a[i:j] \boxplus  b[i:j] \boxplus  \gamma, \label{eq:caryy}  \\
  (a \boxminus b)[i:j] &= a[i:j] \boxminus b[i:j] \boxminus \beta.  \label{eq:borrow}
\end{align}

@@

Let $\mathcal{L}$ denote the euclidean lattice spawned by the rows of the
following $n\times n$ matrix:
\[
  G_{n,k} = \begin{pmatrix}
    1 & a   & a^2 & \dots & a^{n-1} \\
    0 & 2^k & 0   & \dots & 0 \\
    0 & 0   & 2^k & \dots & 0 \\
    \dots & \dots & \dots & \dots & \dots\\
    0 & 0 & 0 & \dots & 2^k \\
  \end{pmatrix}
\]
This lattice contains all $n$-terms geometric progressions of common ratio $a$
modulo $2^k$; therefore the first $n$ terms of the sequence $\LCG_{k}(x, 0)$
give the coordinates of a point of this lattice.

@@

If we can prove that the right side of the inequality is smaller than the first
minimum of the lattice, then we would have proved that $RH$ is indeed the
closest vector we were searching for. Because we have fixed $k = 64$, by
lemma~\ref{thm:close} we have
$\left\| 2^{k-\ell}\widetilde{T} - U' \right\| \leq 2 \sqrt{n}2^{64-\ell}-1
$. So, if we fix $n$ we can search for the minimum number $\ell$ of known
most-significant bits such that:
\[
  \left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-\ell} \leq \lambda_1(\mathcal{L})
\]

@@

But enough fooling around; let's go back to \textsf{PCG64}.

@@

\section{Predicting PCG64 with Unknown Increment}
\label{sec:Cunknown}

@@

\begin{algorithm}
\begin{algorithmic}[1]
  \Function{ReconstructState}{$\Delta S_0, r_0, \dots, r_k, X_0, \dots, X_k$}
  \State \todo{check indices nabla}
  \For{$i = 0, 1, \dots, k$} \Comment{Setup}
  \State $\nabla S_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{128}$
  \State $Y_i \gets X_i \lll r_i$ \Comment{Undo rotations}
  \State $\mathcal{C}_i[0] \gets 0$ \Comment{Bootstrap induction}
  \State $\mathcal{C}_i[64] \gets \bigl( Y_i[0] \oplus Y_i[j] \bigr) \oplus \bigl( \nabla S_i[j] \oplus \nabla S_i[64+j]$
  \EndFor
  \For{$j=1, 2, \dots, 64$} \Comment{Induction}
  \State $\mathfrak{i} \gets \bot$ \Comment{Find good index}
  \For{$k = 1, 2, \dots, k$}
  \If{$\nabla S_{k}[j-1] \neq \mathcal{C}_{k}[j-1] \wedge \nabla S_{k}[64 + j-1] = \mathcal{C}_{k}[64 + j-1]$}
  \State $\mathfrak{i} \gets k$
  \EndIf
  \EndFor
  \If{$\mathfrak{i} = \bot$} \Comment{No suitable indice found?}
  \State \textbf{Abort} with Failure
  \EndIf
  \State  \Comment{Compute next state bit}
  \State $S_0[j-1] \gets Y_0[j-1] \oplus Y_{\mathfrak{i}}[j-1] \oplus \bigl( \nabla S_{\mathfrak{i}}[j-1] \oplus \nabla S_{\mathfrak{i}}[64+j-1] \oplus \nabla S_{\mathfrak{i}}[64 + j-1]\bigr)$
  \State $S_0[64 + j-1] \gets Y_0[64 + j-1] \oplus S_0[j-1]$
  \For{$i = 0, 1, \dots, k$} \Comment{Compute next carries}
  \State $\mathcal{C}_i[j] \gets \MAJ(S_0[j-1], \nabla S_i[j-1], \mathcal{C}_i[j-1])$
  \State $\mathcal{C}_i[64 + j] \gets \MAJ(S_0[64+j-1], \nabla S_i[64+j-1], \mathcal{C}_i[64+j-1])$
  \EndFor
  \EndFor
  \State \Return $S_0$
  \EndFunction
\end{algorithmic}
\caption{Full state reconstruction algorithm}
\label{algo:last}
\end{algorithm}

@@

It actually ran in 35 wall-clock minutes using 512 cluster nodes, each equipped
with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz} (``\emph{Cascade
  Lake}''). The actual machine is the \verb|[----REDACTED FOR PEER REVIEW----]|
computer located at the \verb|[----ALSO REDACTED----]| computation center. Note
that on this particular parallel computer, running the algorithm with $\ell=13$
would take 10 hours using the same amount of resources, so the whole procedure
\emph{is} practical, even in the absolute worst case.

@@

\section{Conclusion}
\todo{y'a pas de conclusion}
% pas dans la version anonyme.
% \paragraph{Acknowledgements.} The authors thank Damien Vergnaud and Vincent
% Zucca for helpful discussions. We also thank Léo Ducas for helping with the
% \texttt{G6K} library.

@@

%\appendix

>>>>

\section{Introduction}

@@

@@

Not all pseudo-random generators are of cryptographic strength. In some
applications, they are simply not necessary: to be used in Monte-Carlo
numerical simulations or generate random choices in games, a relaxed,
non-cryptographic notion of pseudo-randomness may be sufficient. This allows for
faster algorithms. For instance, \textsf{python} standard library's
\texttt{random} module uses the Mersenne Twister~\cite{MatsumotoN98}. The
\textsf{C} library that comes along \texttt{gcc} (the \texttt{glibc}) uses a
(poor) truncated linear congruential generator by default to implement the
\texttt{rand} function.

@@

In any case, designers of conventional pseudo-random generators try to obtain
the simplest and fastest algorithm that passes the day's favourite test suite. A
few selected ones are shown in Fig.~\ref{fig:prngs}. \texttt{lehmer64} is a
truncated linear congruential generator, touted as ``the fastest PRNG that
passes \texttt{BigCrush}''~\cite{LemireLehmer}. \texttt{xorshift128} is a clever
implementation of a 128-bit LFSR with period $2^{128} - 1$ due to
Marsaglia~\cite{Marsaglia03}, using only a few simple 32-bit
operations. \texttt{xorshift128+} is a improved version due to
Vigna~\cite{Vigna17} that returns the sum of two consecutive outputs of a
Xorshift LFSR; it passes the ``BigCrush'' test suite and is the default PRNG in
many Javascript implementations, including that in Google's V8 engine (Chrome),
Firefox and Safari.

@@

\textsf{PCG} stands for ``Permuted Congruential Generator'': it essentially
consists in applying a non-linear filtering function on top of a linear
congruential generator (in a way reminiscent to the venerable filtered
LFSRs). The resulting combination is fast and passes current statistical test
suites. The \textsf{PCG} family contains many members, but we focus on the
strongest one, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has a
128-bit internal state and produces 64 bits when clocked. It is the default
pseudo-random number generator in the popular \textsf{NumPy}~\cite{Numpy}
scientific computing package for \textsf{Python}.

@@

When the increment is unknown, things are a bit more complicated. This is the
default situation in \textsf{NumPy}, where both the state and the increment are
initialised using an external source of entropy. In this case, our prediction
algorithm requires 4096 bits of pseudo-random stream ; it guesses between 51 and
55 bits, then for each guess it solves an instance of CVP in dimension 4 (using
about 75 arithmetic operations). This recovers 64 more bits of information about
the difference between two successive states, and this is enough to filter the
bad guesses. This information can then be used in a subsequent and comparably
inexpensive phase to recover the entire internal state. On average, the whole
process requires a bit less than 20 000 CPU hours to complete.

We implemented the algorithms in \textsf{C}, then asked the designer of the
\textsf{PCG} family to send us ``challenge'' pseudo-random streams ; we ran our
code and emailed back the (correct) seeds used to generate the challenge streams the next
day.

@@

Knuth discussed whether \emph{truncated} linear congruential generators could be
good stream ciphers; he therefore studied the problem of recovering the internal
state of a truncated linear congruential generator
$x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$ are unknown~\cite{Knuth85}; he
gave an algorithm exponential in the number of truncated bits.

Boyar studied further the problem~\cite{Boyar1989} and presented an algorithm
which could predict a linear congruential generator when all the parameters
(multiplier, increment, modulus and initial state) are unknown; she extended her
idea to the case of truncated linear congruential generators, under the
condition that the number of bit unrevealed is really small in comparison to the
size of the modulus.

Frieze \textit{et al}~\cite{Frieze} improved the efficiency of reconstruction
algorithms in simpler cases. They supposed that the multiplier $a$ and the
modulus $2^k$ were known and used lattice-based techniques to recover a
truncated linear congruential generator more efficiently, with more truncated
bits.  more efficiently.

Later on, Joux and Stern extended this result to the case where the multiplier
$a$ and the modulus $2^k$ are unknown, also using lattice
techniques~\cite{JouxS98}.

@@

\textsf{PCG64} has an internal state of 128-bit, which operate as a linear
congruential generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]
Where the ``multiplier'' $a$ is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
the generator. The increment $c$ may be specified by the user of the PRNG to
produce different output streams with the same seed (just as the IV acts in a
stream cipher). If no value of $c$ is specified, then a default increment is
provided. Note that $c$ must be odd. The default values are:
\begin{align*}
  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
\end{align*}

@@

In the rest of this paper, we often perform arithmetic operations on integers
where only some bits are known. This leads to generation of unknown carries. If
$a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there is a
\emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
$0 \leq \beta \leq 1$) such that:
\begin{align}
  (a \boxplus b)[i:j]  &= a[i:j] \boxplus  b[i:j] \boxplus  \gamma, \label{eq:caryy}  \\
  (a \boxminus b)[i:j] &= a[i:j] \boxminus b[i:j] \boxminus \beta.  \label{eq:borrow}
\end{align}

@@

Let $\mathcal{L}$ denote the euclidean lattice spawned by the rows of the
following $n\times n$ matrix:
\[
  G_{n,k} = \begin{pmatrix}
    1 & a   & a^2 & \dots & a^{n-1} \\
    0 & 2^k & 0   & \dots & 0 \\
    0 & 0   & 2^k & \dots & 0 \\
    \dots & \dots & \dots & \dots & \dots\\
    0 & 0 & 0 & \dots & 2^k \\
  \end{pmatrix}
\]
This lattice contains all $n$-terms geometric progressions of common ratio $a$
modulo $2^k$; therefore the first $n$ terms of the sequence $\LCG_{k}(x, 0)$
give the coordinates of a vector in this lattice.

@@

Note that $\vvvert H^{-1} \vvvert \times \vvvert H \vvvert$ is the condition
number of the matrix $H$. Lattice reduction has the side effect of reducing the
condition number, therefore it makes sense to use an LLL-reduced basis. If we
can prove that the right side of the inequality is smaller than the first
minimum of the lattice, then we would have proved that $RH$ is indeed the
closest vector we were searching for. Because we have fixed $k = 64$, by
lemma~\ref{thm:close} we have
$\left\| 2^{k-\ell}\widetilde{T} - U' \right\| \leq 2 \sqrt{n}2^{64-\ell}-1
$. So, if we fix $n$ we can search for the minimum number $\ell$ of known
most-significant bits such that:
\[
  \left(1 + \vvvert H \vvvert \times \left\vvvert H^{-1} \right\vvvert \right) 2 \sqrt{n} 2^{64-\ell} \leq \lambda_1(\mathcal{L})
\]

@@

%\noindent But enough fooling around; let's go back to \textsf{PCG64}.

@@

\section{State Reconstruction for \textsf{PCG64} With Secret Increment}
\label{sec:Cunknown}

@@

\begin{algorithm}
\begin{algorithmic}[1]
  \Function{ReconstructState}{$\Delta S_0, r_0, \dots, r_k, X_0, \dots, X_k$}
  \For{$i = 0, 1, \dots, k$} \Comment{Setup}
  \State $\nabla S_i \gets \Delta S_0 (a^i-1)(a-1)^{-1} \bmod 2^{128}$
  \State $Y_i \gets X_i \lll r_i$ \Comment{Undo rotations}
  \State $\mathcal{C}_i[0] \gets 0$ \Comment{Bootstrap induction}
  \State $\mathcal{C}_i[64] \gets \bigl( Y_i[0] \oplus Y_i[j] \bigr) \oplus \bigl( \nabla S_i[j] \oplus \nabla S_i[64+j]$
  \EndFor
  \For{$j=1, 2, \dots, 64$} \Comment{Induction}
  \State $\mathfrak{i} \gets \bot$ \Comment{Find good index}
  \For{$k = 1, 2, \dots, k$}
  \If{$\nabla S_{k}[j-1] \neq \mathcal{C}_{k}[j-1] \wedge \nabla S_{k}[64 + j-1] = \mathcal{C}_{k}[64 + j-1]$}
  \State $\mathfrak{i} \gets k$
  \EndIf
  \EndFor
  \If{$\mathfrak{i} = \bot$} \Comment{No suitable indice found?}
  \State \textbf{Abort} with Failure
  \EndIf
  \State  \Comment{Compute next state bit}
  \State $S_0[j-1] \gets Y_0[j-1] \oplus Y_{\mathfrak{i}}[j-1] \oplus \bigl( \nabla S_{\mathfrak{i}}[j-1] \oplus \nabla S_{\mathfrak{i}}[64+j-1] \oplus \nabla S_{\mathfrak{i}}[64 + j-1]\bigr)$
  \State $S_0[64 + j-1] \gets Y_0[64 + j-1] \oplus S_0[j-1]$
  \For{$i = 0, 1, \dots, k$} \Comment{Compute next carries}
  \State $\mathcal{C}_i[j] \gets \MAJ(S_0[j-1], \nabla S_i[j-1], \mathcal{C}_i[j-1])$
  \State $\mathcal{C}_i[64 + j] \gets \MAJ(S_0[64+j-1], \nabla S_i[64+j-1], \mathcal{C}_i[64+j-1])$
  \EndFor
  \EndFor
  \State \Return $S_0$
  \EndFunction
\end{algorithmic}
\caption{Full state reconstruction algorithm}
\label{algo:last}
\end{algorithm}

@@

It actually ran in 35 wall-clock minutes using 512 cluster nodes, each equipped
with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz} (``\emph{Cascade
  Lake}''). The actual machine is the \verb|[----REDACTED----]| computer located
at the \verb|[----ALSO REDACTED----]| computation center. Note that on this
particular parallel computer, running the algorithm with $\ell=13$ would take 10
hours using the same amount of resources, so the whole procedure \emph{is}
practical, even in the absolute worst case.

@@

\section{Conclusion}

We have presented a practical state reconstruction algorithm for the
\textsf{PCG64} conventional pseudo-random number generator, the default in the
\texttt{NumPy} library. In the worst case, we recovers all the secret
information using 512 consecutive output bytes, using 2.3 CPU years of
computation. We have executed the algorithm in practice using a large parallel
computer. The \textsf{PCG64} generator is fast and not intended for
cryptographic purposes; we have shown that, in practice, this comes at the price
of strong pseudo-randomness. It should absolutely not be used when
unpredictability of the random numbers is required, for fear of practical
attacks.

On the other hand, our results do not mean that \textsf{PCG64} should be
deprecated for scientific computing. But they do mean that its output has
detectable properties. Whether these properties may affect the results of
Monte-Carlo numerical simulations is another matter entirely.

% pas dans la version anonyme.
% \paragraph{Acknowledgements.} The authors thank Damien Vergnaud and Vincent
% Zucca for helpful discussions. We also thank Léo Ducas for helping with the
% \texttt{G6K} library.

@@
